Modeling Ionic Polymer-Metal Composites\\ with Space-Time Adaptive Multimesh \emph{hp}-FEM

### 2010-10-19 conclusions.tex
Work on the first two pages.
Results and conclusions improved

\section{Conclusions}

In this work the system of Nernst-Planck-Poisson equations
was solved using \emph{hp}-finite element method with adaptive
multimesh configuration. The weak form, residuals and
the Jacobian matrix of the system were explicitly derived
and implemented in Hermes \emph{hp}-FEM adaptive solver.
The solution for Nernst-Planck-Poisson
problem with two field variables $C$ and $\phi$ results in 
very different field gradients in the space and time.
Conventional finite element solvers do not provide the means 
how to deal with such problems such that both the error of
the solution and problem size remain small throughout the
time dependent solving process.

It was shown that using time dependent adaptivity, multi-mesh
configuration, and anisotropic \emph{hp} refinements, the problem
size remains acceptably small throughout the solving process.
Namely, hermes refinement modes HP\_ANISO and HP\_ANISO\_P
resulted the smallest and fastest problem, respectively.
At the same time, relative error of the solution was known.
Furthermore, using the multi-mesh configuration for the variables
$C$ and $\phi$ was justified --- the adaptivity algorithm
did not refine the mesh of $\phi$ nor did increase the
polynomial degree throughout the adaptivity process. However,
the mesh was significantly refined for $C$ and also the
maximum polynomial degree was varied in the range of
$2\ldots 9$. So it is efficient to use multi-mesh in terms of
the number of degrees of freedom.

Conclusively, by using \emph{hp}-FEM with adaptive multi-mesh
configuration we can possibly reduce the problem size
of the Nernst-Planck-Poisson equation system significantly while
still maintaining prescribed precision of the solution. 
We believe, and this
is yet to be demonstrated, that this becomes especially
important when dealing with 3D problems with a large physical
domain.

>>>>

\section{Conclusion and Outlook}\label{sec:conc}

In this work the system of Nernst-Planck-Poisson equations
was solved using \emph{hp}-finite element method with adaptive
multimesh configuration. The weak form, residuals and
the Jacobian matrix of the system were explicitly derived
and implemented in Hermes \emph{hp}-FEM adaptive solver.
The solution for Nernst-Planck-Poisson
problem with two field variables $C$ and $\phi$ results in 
very different field gradients in the space and time.
Conventional finite element solvers do not necessarily provide the means 
how to deal with such problems such that both the error of
the solution and problem size remain small throughout the
time dependent solving process. This is largely due to the
fact that it is very difficult to find an optimal initial mesh.

It was shown that using time dependent adaptivity, multi-mesh
configuration, and anisotropic \emph{hp} refinements, the problem
size remains acceptably small throughout the solving process.
Namely, Hermes refinement modes HP\_ANISO and HP\_ANISO\_P
resulted the smallest and fastest problem, respectively.
At the same time, relative error of the solution was known.
Furthermore, using the multi-mesh configuration for the variables
$C$ and $\phi$ was justified --- the adaptivity algorithm
did not refine the mesh of $\phi$ nor did increase the
polynomial degree throughout the adaptivity process. However,
the mesh was significantly refined for $C$ and also the
maximum polynomial degree was varied in the range of
$2\ldots 9$. So it is efficient to use multi-mesh in terms of
the number of degrees of freedom.

Conclusively, by using \emph{hp}-FEM with adaptive multi-mesh
configuration we can possibly reduce the problem size
of the Nernst-Planck-Poisson equation system significantly while
still maintaining prescribed precision of the solution. 
We believe, and this
is yet to be demonstrated, that this becomes especially
important when dealing with 3D problems in a large physical
domain with non-uniform boundary conditions.

### 2010-10-19 model.tex
Results and conclusions improved

Hermes implements a novel adaptive multimesh $hp$-FEM [ADD REFERENCES]
that makes it possible to approximate different fields on individual meshes,
without breaking the monolithic structure of the coupling mechanism.
For practical reasons, the meshes in the system are not allowed to be 
completely independent -- they have a common coarse mesh that we call {\em 
master mesh}. The master mesh is 
there for algorithmic purposes only and it may not even be used for 
discretization purposes. Every mesh in the system is obtained from 
the master mesh via an arbitrary sequence of elementary refinements. 
Assembling is done on a {\em union mesh}, a geometrical union of 
all meshes in the system (imagine printing all meshes on transparencies and 
positioning them on top of each other).

>>>>

Hermes implements a novel adaptive multimesh $hp$-FEM~\cite{solin2010monolithic,
solin2010adaptive,dubcova2010space}
that makes it possible to approximate different fields on individual meshes,
without breaking the monolithic structure of the coupling mechanism.
For practical reasons, the meshes in the system are not allowed to be 
completely independent -- they have a common coarse mesh that we call {\em 
master mesh}. The master mesh is 
there for algorithmic purposes only and it may not even be used for 
discretization purposes. Every mesh in the system is obtained from 
the master mesh via an arbitrary sequence of elementary refinements. 
Assembling is done on a {\em union mesh}, a geometrical union of 
all meshes in the system (imagine printing all meshes on transparencies and 
positioning them on top of each other).

### 2010-10-19 results.tex
Results and conclusions improved

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=.8\columnwidth]{mesh}
  \caption{\label{fig:mesh} Initial coarse mesh (a),
  	half refined mesh (b) and refined mesh (c). The coarse mesh
	and refined mesh were used in the initial calculations, the latter one
	in case of \emph{p}-adaptivity (including HP\_ANISO\_P). The half-refined mesh was
	used later to optimize the \emph{hp}-adaptive refinement solutions.}
  \end{centering}
\end{figure}
Calculation for all the refinement modes were performed
in both single-mesh and multi-mesh configurations. 
The following numerical results were recorded for each 
refinement mode: converged relative error, cumulative CPU
time, and the problem size in terms of number
of degrees of freedoms (NDOFs) at each time step. 
Two types of initial meshes were used --- in case of only \emph{p}-adaptivity,
more refined mesh was used (Fig.~\ref{fig:mesh}~(c)) to ensure
the error convergence.
When the element size refinement
was also enabled (all \emph{h}/\emph{hp} refinement modes), very coarse initial mesh
was used (see Fig.~\ref{fig:mesh}~(a)) to let the adaptivity
algorithm find the most optimal mesh. So coarse initial mesh might not be
suitable for most the practical applications, as will be demonstrated in the
end of this section, however, it provides a good insight into the
adaptivity performance of Hermes.
In both cases, the initial mesh was loaded at each time step of the
calculation.

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=.75\columnwidth]{cphi}
  \caption{\label{fig:cphi} Concentration $C$
  and voltage $\phi$ at two different time steps
  (HP\_ANISO refinement mode was used). This is a 2D solution shown
  as in 3D, where the height indicates the values of $C$ and $\phi$.}
  \end{centering}
\end{figure}
Example of the solution at $t=0.7\ s$ and $t=3.0\ s$ 
calculated with HP\_ANISO refinement mode is shown
in Fig.~\ref{fig:cphi}. The time $t=0.7\ s$ was chosen because
by that time (with given constants in Table~\ref{Table:used-constants}),
some ionic migration has already taken place and
the concentration gradients near the boundaries $\partial\Omega_1$ and
$\partial\Omega_3$ have formed. The automatic mesh refinements
at different time steps are clearly visible in the figure --- especially
near $\partial\Omega_1$, where the $C_{\Omega_1}\approx 0$ and
the gradient $\nabla C$ is
moving in time (this was also demonstrated in Fig.~\ref{fig:comsol-conc-volt}).

@@

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=\columnwidth]{singlemulti_dof}
  \caption{\label{fig:singlemultidof} NDOFs in case 
  of single-mesh and multi-mesh solutions with HP\_ISO
  and HP\_ANISO refinement modes. (Notice log Y scale)}
  \end{centering}
\end{figure}

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=\columnwidth]{singlemulti_cpu}
  \caption{\label{fig:singlemulticpu} CPU times in case
  of single-mesh and multi-mesh solutions with HP\_ISO
  and HP\_ANISO refinement modes. (Notice log Y scale)}
  \end{centering}
\end{figure}

@@

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=\columnwidth]{isoaniso_cpu}
  \caption{\label{fig:isoanisocpu} CPU times in case 
  of multi-mesh configuration with H\_ISO, H\_ANISO,
  HP\_ISO, HP\_ANISO, and single-mesh configuration with HP\_ANISO\_H
  refinement modes. (notice log Y scale)}
  \end{centering}
\end{figure}

@@

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=\columnwidth]{isoanisop_cpu}
  \caption{\label{fig:isoanisopcpu} CPU times in case 
  of multi-mesh configuration with P\_ISO, P\_ANISO, and
  HP\_ANISO\_P refinement modes.}
  \end{centering}
\end{figure}
To narrow down the list of refinement modes for the given problem, first the
isotropic and anisotropic adaptivities were compared. The \emph{p}-adaptivity
modes were compared separately as they used more refined initial mesh 
(shown in Fig.~\ref{fig:mesh}~(c)).
Fig.~\ref{fig:isoanisodof} and Fig.~\ref{fig:isoanisocpu} provide the comparison
of H\_ISO, H\_ANISO, HP\_ISO, HP\_ANISO, and HP\_ANISO\_H modes in terms of CPU time
and NDOFs.
Fig.~\ref{fig:isoanisopdof} and Fig.~\ref{fig:isoanisopcpu} show the similar
comparsion for the P\_ISO, P\_ANISO, and HP\_ANISO\_P modes.
It is not difficult to see that the anisotropic refinement modes result in a reasonable problem
size and the problems solve within a reasonable calculation time. It is interesting
to note that HP\_ANISO results in the smallest problem size. Also, in
\emph{p}-adaptivity group, HP\_ANISO\_P results in the smallest problem size
at each time step, whereas P\_ISO and P\_ANISO result in a very large problem size
during the first time steps of the solution.
Here the term ``reasonable problem size''
means that the number of degrees of freedom in time converges
to so that $N_{DOF}<500$, and the term ``reasonable calculation time''
means that the calculation (step $\tau=0.01\ s$, physical
time $t_{end}=3.0\ s$) time $t$ on a givem system was $t<500\ s$.
Although these parameters are empirical, they serve as an upper limit, given
that the most refinement modes give significantly better results, i.e.
$t<<500\ s$ and $N_{DOF} << 500$.

@@

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=\columnwidth]{cpu}
  \caption{\label{fig:cpu} Comparative CPU time for different refinement modes.}
  \end{centering}
\end{figure}

@@

When it comes to a large domain or 3D modeling, the problem
size becomes a very important factor. Namely, 3D full scale solutions tend to use
a lot of memory. Therefore, we consider HP\_ANISO the most
suitable refinement mode for the given problem. 
Hence, some ways to optimize the model based on HP\_ANISO
mode to improve the CPU time
%TODO, continue here
factor without significantly compromising the NDOFs will be considered. 
The desirable output would be HP\_ANISO problem size
close to HP\_ANISO\_P CPU time. One way to optimize the problem
is to choose more refined initial mesh. Another way would be to change
the refinement frequency during the solving process. In theory, this
could result in a fewer adaptivity steps which are quite expensive in terms
of CPU time. Recall that up to this point,
the initial mesh has been loaded in the beginning of each time step. 
True, by employing these optimizations, we must know something about the problem
and its solution beforehand. However, it could still be practial when solving a real problem
in a large domain.

The problem size and CPU time with HP\_ANISO
and HP\_ANISO\_P adaptivities on more refined initial mesh (see Fig.~\ref{fig:mesh}~(c))
compared to the coarse initial mesh (Fig.~\ref{fig:mesh}~(a)) and 
HP\_ANISO\_P solution is shown in Fig.~\ref{fig:refined-cpu} and Fig.~\ref{fig:refined-dof}.
By using initially more refined mesh, the problem solving time
can be reduced in case of HP\_ANISO, at the same time, the problem size increases.
In this situation, HP\_ANISO\_P and HP\_ANISO perform equally well.

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=\columnwidth]{unreffreq_dof}
  \caption{\label{fig:unreffreq-dof} NDOFs at each time step for
  HP\_ANISO and HP\_ANISO\_P with mesh unrefinement at each time step and at over each
  	time step after first $0.5\ s$ physical solution time.}
  \end{centering}
\end{figure}
The next proposed optimization involved changing the unrefinement frequency.
It is known that the concentration gradient $\nabla C$ changes the most in the initial phase
of the calculation, therefore, the unrefinement after each time step was performed
until $t=0.5\ s$ (physical time). After that, the
unrefinement was performed in $\Delta t = 0.10\ s$ interval.
However, this optimization did not result in a stable solution, i.e. the problem size 
does not remain steady, but starts to oscillate depending on the unrefinement
frequency. This is shown in 
Fig.~\ref{fig:unreffreq-dof}.

Therefore varying the unrefinement frequency will
not likely result in desired results in real applications for given system of equation.
At the same time, by varying a mesh size, optimal initial mesh could be found
for both HP\_ANISO and HP\_ANISO\_P refinement modes.

>>>>

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=.8\columnwidth]{mesh}
  \caption{\label{fig:mesh} Initial coarse mesh (a),
  	half refined mesh (b) and refined mesh (c). The coarse mesh (a)
	and refined mesh (c) were used in the initial calculations, the latter one
	in case of \emph{p}-adaptivity (including HP\_ANISO\_P). The half-refined mesh was
	used later to optimize the \emph{hp}-adaptive refinement solutions.}
  \end{centering}
\end{figure}
Calculation for all the refinement modes were performed
in both single-mesh and multi-mesh configurations. 
The following numerical results were recorded for each 
refinement mode: converged relative error, cumulative CPU
time, and the problem size in terms of number
of degrees of freedoms (NDOFs) at each time step. 
Two types of initial meshes were used --- in case of only \emph{p}-adaptivity,
more refined mesh was used (Fig.~\ref{fig:mesh}~(c)) to ensure
the error convergence.
When the element size refinement
was also enabled (all \emph{h}/\emph{hp} refinement modes), very coarse initial mesh
was used (see Fig.~\ref{fig:mesh}~(a)) to let the adaptivity
algorithm find the most optimal mesh. That coarse initial mesh might not be
suitable for practical applications, as will be demonstrated in the
end of this section, however, it provides a good insight into the
adaptivity performance of Hermes.
In both cases, the initial mesh was loaded at each time step of the
calculation.

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=.75\columnwidth]{cphi}
  \caption{\label{fig:cphi} Concentration $C$
  and voltage $\phi$ at two different time steps
  (HP\_ANISO refinement mode was used). This is a 2D solution shown
  in 3D, where the height indicates the values of $C$ and $\phi$.}
  \end{centering}
\end{figure}
Example of the solution at $t=0.7\ s$ and $t=3.0\ s$ 
calculated with HP\_ANISO refinement mode is shown
in Fig.~\ref{fig:cphi}. The time $t=0.7\ s$ was chosen because
by that time (with given constants in Table~\ref{Table:used-constants}),
some ionic migration has already taken place and
the concentration gradients near the boundaries $\partial\Omega_1$ and
$\partial\Omega_3$ have formed. The automatic mesh refinements
at different time steps are clearly visible in the figure --- especially
near $\partial\Omega_1$, where the $C_{\Omega_1}\approx 0$ and
the gradient $\nabla C$ is
shifting in time (this is also demonstrated in Fig.~\ref{fig:comsol-conc-volt}).

@@

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=\columnwidth]{singlemulti_dof}
  \caption{\label{fig:singlemultidof} NDOFs in case 
  of single-mesh and multi-mesh configurations with HP\_ISO
  and HP\_ANISO refinement modes. (Notice log Y scale)}
  \end{centering}
\end{figure}

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=\columnwidth]{singlemulti_cpu}
  \caption{\label{fig:singlemulticpu} Cumulative CPU times in case
  of single-mesh and multi-mesh configurations with HP\_ISO
  and HP\_ANISO refinement modes. (Notice log Y scale)}
  \end{centering}
\end{figure}

@@

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=\columnwidth]{isoaniso_cpu}
  \caption{\label{fig:isoanisocpu} Cumulative CPU times in case 
  of multi-mesh configuration with H\_ISO, H\_ANISO,
  HP\_ISO, HP\_ANISO, and single-mesh configuration with HP\_ANISO\_H
  refinement modes. (notice log Y scale)}
  \end{centering}
\end{figure}

@@

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=\columnwidth]{isoanisop_cpu}
  \caption{\label{fig:isoanisopcpu} Cumulative CPU times in case 
  of multi-mesh configuration with P\_ISO, P\_ANISO, and
  HP\_ANISO\_P refinement modes.}
  \end{centering}
\end{figure}
To narrow down the list of refinement modes for the given problem, first the
isotropic and anisotropic adaptivities were compared. The \emph{p}-adaptivity
modes were compared separately as they used more refined initial mesh 
(shown in Fig.~\ref{fig:mesh}~(c)).
Fig.~\ref{fig:isoanisodof} and Fig.~\ref{fig:isoanisocpu} provide the comparison
of H\_ISO, H\_ANISO, HP\_ISO, HP\_ANISO, and HP\_ANISO\_H modes in terms of 
cumulative CPU time and NDOFs.
Fig.~\ref{fig:isoanisopdof} and Fig.~\ref{fig:isoanisopcpu} show the similar
comparsion for the P\_ISO, P\_ANISO, and HP\_ANISO\_P modes.
It is not difficult to see that the anisotropic refinement modes result in a reasonable problem
size and the problems solve within a reasonable calculation time. It is interesting
to note that HP\_ANISO results in the smallest problem size. Also, in
\emph{p}-adaptivity group, HP\_ANISO\_P results in the smallest problem size
at each time step, whereas P\_ISO and P\_ANISO result in a very large problem size
during the first time steps of the solution.
Here the term ``reasonable problem size''
means that the number of degrees of freedom in time converges
to so that $N_{DOF}<500$, and the term ``reasonable calculation time''
means that the calculation (step $\tau=0.01\ s$, physical
time $t_{end}=3.0\ s$) time $t$ on a givem system was $t<500\ s$.
Although these parameters are empirical, they serve as an upper limit, given
that the most refinement modes give significantly better results ---
$t<<500\ s$ and $N_{DOF} << 500$.

@@

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=\columnwidth]{cpu}
  \caption{\label{fig:cpu} Comparative cumulative CPU time for different refinement modes.}
  \end{centering}
\end{figure}

@@

When it comes to a large domain or 3D modeling, the problem
size becomes a very important factor. Namely, 3D full scale solutions tend to use
a lot of memory. Therefore, we consider HP\_ANISO the most
suitable refinement mode for the given problem. 
Hence, some ways to optimize the HP\_ANISO model
to improve the CPU time factor without significantly 
compromising the NDOFs will be considered. 
The desirable output would be HP\_ANISO problem size
close to HP\_ANISO\_P CPU time. 
One way to optimize the problem
is to choose more refined initial mesh. Another way would be to change
the refinement frequency during the solving process. Theoretically it 
could result in a fewer adaptivity steps which are rather expensive in terms
of CPU time. Recall that up to this point,
the initial mesh was loaded in the beginning of each time step. 
True, by employing the optimizations, one must know something about the problem
and its solution beforehand. However, it could still be practial when solving a real problem
in a large domain.

The problem size and CPU time with HP\_ANISO
and HP\_ANISO\_P adaptivities on more refined initial mesh (see Fig.~\ref{fig:mesh}~(c))
compared to the coarse initial mesh (Fig.~\ref{fig:mesh}~(a)) and 
HP\_ANISO\_P solution are shown in Fig.~\ref{fig:refined-cpu} and Fig.~\ref{fig:refined-dof}.
By using initially more refined mesh, the problem solving time
can be reduced in case of HP\_ANISO, at the same time, the problem size increases
as the initial mesh is not necessarily the most optimal one.
However, in this situation, HP\_ANISO\_P and HP\_ANISO perform equally well.

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=\columnwidth]{unreffreq_dof}
  \caption{\label{fig:unreffreq-dof} NDOFs at each time step for
  HP\_ANISO and HP\_ANISO\_P with mesh unrefinement at each time step and at over each
  	time step after first $0.5\ s$ physical solution time.}
  \end{centering}
\end{figure}
The next proposed optimization involved changing the unrefinement frequency.
It is known that the concentration gradient $\nabla C$ changes the most in the initial phase
of the calculation, therefore, the unrefinement after each time step was performed
until $t=0.5\ s$ (physical time). After that, the
unrefinement was performed in $\Delta t = 0.10\ s$ interval.
However, this optimization does not appear to result in a stable solution, 
i.e. the problem size 
does not remain steady, but starts to oscillate depending on the unrefinement
frequency. This is shown in Fig.~\ref{fig:unreffreq-dof}.

Therefore varying the unrefinement frequency will
likely not result in desired results in real applications for given system of equation.
At the same time, by varying the initial mesh size, optimal initial mesh could be found
for both HP\_ANISO and HP\_ANISO\_P refinement modes.
